---
title: "Clustering Made Simple in R: One Function to Rule Them All"
description: |
  Learn how to build a generic R function that can apply multiple clustering algorithms (K-means, hierarchical clustering, DBSCAN, HDBSCAN, PAM) and automatically visualize results using PCA or UMAP. A practical, reusable tool designed for biologists and data analysts working with tabular data (imaging, omics, cytometry, etc.).
author:
  - name: Cédric Hassen-Khodja
    url: {}
date: 2025-12-05
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c(
  "palmerpenguins",
  "dplyr",
  "ggplot2",
  "dbscan",
  "uwot",
  "tibble",
  "cluster"
)

new_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(new_packages) > 0) install.packages(new_packages)

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(dbscan)
library(uwot)
library(tibble)
library(cluster)

```

## Why a generic clustering function?

If you work with **tabular data** (high-content imaging, omics, cytometry, single-cell, etc.), you’ve probably repeated this workflow many times:

1. Pick a clustering method (K-means, HDBSCAN, DBSCAN, hierarchical, …)
2. Standardize features
3. Run the clustering
4. Apply a dimensionality reduction (PCA, UMAP)
5. Plot points in 2D, colored by cluster
6. Repeat all of this when you switch method or dataset

This is:

* **Repetitive** (lots of copy-paste)
* **Error-prone** (easy to forget scaling or change parameters)
* Not very **reusable** across projects

In this post, we’ll build **one generic R function** that:

* Accepts a data frame (rows = objects, columns = features)
* Supports several clustering methods:

  * `"kmeans"`
  * `"hclust"` (hierarchical clustering)
  * `"dbscan"`
  * `"hdbscan"`
  * `"pam"` (partitioning around medoids)
* Uses either `"PCA"` or `"UMAP"` for 2D visualization
* Returns:

  * The **original data with cluster labels**
  * The **clustering model**
  * The **2D embedding**
  * A ready-to-use **`ggplot`**

We’ll demonstrate it on the **Palmer Penguins** dataset.

## Data preparation

```{r data-prep}
data("penguins", package = "palmerpenguins")

penguins_clean <- penguins %>%
  filter(
    !is.na(bill_length_mm),
    !is.na(bill_depth_mm),
    !is.na(flipper_length_mm),
    !is.na(body_mass_g)
  )

penguins_features <- penguins_clean %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

head(penguins_features)
```

This is the kind of numeric table you would also get from:

* Imaging analysis (per-cell or per-nucleus features)
* Proteomics / transcriptomics (per-sample features)
* Cytometry / single-cell pipelines

## A generic `cluster_and_embed()` function

Here is the core function. It:

* Keeps **non-numeric columns** as metadata
* Uses only **numeric columns** for clustering
* Scales features (optional)
* Applies **PCA** or **UMAP**
* Runs the selected clustering method
* Builds a 2D scatter plot colored by cluster

```{r cluster-and-embed-fun}
cluster_and_embed <- function(
  data,
  method = c("kmeans", "hclust", "dbscan", "hdbscan", "pam"),
  dimred = c("PCA", "UMAP"),
  k = NULL,                # for kmeans / hclust / pam
  hclust_method = "ward.D2",
  eps = NULL,              # for DBSCAN
  minPts = 5,              # for DBSCAN / HDBSCAN
  scale = TRUE,
  seed = 123,
  umap_n_neighbors = 15,
  umap_min_dist = 0.1
) {
  method <- match.arg(method)
  dimred <- match.arg(dimred)
  
  # Ensure data frame
  data_input <- as.data.frame(data)
  
  # Keep numeric columns for clustering
  num_cols <- sapply(data_input, is.numeric)
  if (!all(num_cols)) {
    warning(
      "Non-numeric columns removed for clustering: ",
      paste(names(data_input)[!num_cols], collapse = ", ")
    )
  }
  data_numeric <- data_input[, num_cols, drop = FALSE]
  
  if (ncol(data_numeric) == 0) {
    stop("No numeric columns available for clustering.")
  }
  
  mat <- as.matrix(data_numeric)
  if (scale) {
    mat <- scale(mat)
  }
  
  set.seed(seed)
  
  # ---- Dimensionality reduction ----
  if (dimred == "PCA") {
    dr <- prcomp(mat)
    embed <- as.data.frame(dr$x[, 1:2, drop = FALSE])
    colnames(embed) <- c("Dim1", "Dim2")
    xlab <- "PC1"
    ylab <- "PC2"
  } else { # UMAP
    embed_mat <- uwot::umap(
      mat,
      n_neighbors = umap_n_neighbors,
      min_dist = umap_min_dist,
      n_components = 2,
      verbose = FALSE,
      seed = seed
    )
    embed <- as.data.frame(embed_mat)
    colnames(embed) <- c("Dim1", "Dim2")
    xlab <- "UMAP1"
    ylab <- "UMAP2"
  }
  
  # ---- Clustering ----
  if (method == "kmeans") {
    if (is.null(k)) {
      stop("Please provide 'k' (number of clusters) for k-means.")
    }
    model <- stats::kmeans(mat, centers = k)
    cluster <- factor(model$cluster)
    
  } else if (method == "hclust") {
    if (is.null(k)) {
      stop("Please provide 'k' (number of clusters) for hierarchical clustering.")
    }
    hc <- stats::hclust(dist(mat), method = hclust_method)
    cl_vec <- stats::cutree(hc, k = k)
    model <- hc
    cluster <- factor(cl_vec)
    
  } else if (method == "dbscan") {
    if (is.null(eps)) {
      stop("Please provide 'eps' for DBSCAN.")
    }
    model <- dbscan::dbscan(mat, eps = eps, minPts = minPts)
    cluster <- factor(model$cluster)  # 0 = noise
    
  } else if (method == "hdbscan") {
    model <- dbscan::hdbscan(mat, minPts = minPts)
    cluster <- factor(model$cluster)  # 0 = noise
    
  } else if (method == "pam") {
    if (is.null(k)) {
      stop("Please provide 'k' (number of clusters) for PAM.")
    }
    model <- cluster::pam(mat, k = k)
    cluster <- factor(model$clustering)
  }
  
  # Add cluster to embedding
  embed$cluster <- cluster
  
  # ---- Plot ----
  p <- ggplot(embed, aes(x = Dim1, y = Dim2, color = cluster)) +
    geom_point(alpha = 0.7, size = 2) +
    labs(
      title = paste0("Clustering with ", toupper(method), " (", dimred, ")"),
      x = xlab,
      y = ylab,
      color = "Cluster"
    ) +
    theme_minimal()
  
  # ---- Return all useful objects ----
  result <- list(
    data      = dplyr::bind_cols(data_input, cluster = cluster),
    model     = model,
    embedding = embed,
    plot      = p,
    method    = method,
    dimred    = dimred
  )
  
  return(result)
}
```

This is your **“one function to rule them all”** for standard clustering + visualization on numeric tables.

## Example 1 – K-means + PCA on Penguins

We know there are 3 species in the dataset, so let’s ask K-means for **3 clusters** and visualize using PCA.

```{r kmeans-pca}
res_kmeans_pca <- cluster_and_embed(
  data   = penguins_features,
  method = "kmeans",
  dimred = "PCA",
  k      = 3,
  scale  = TRUE
)

res_kmeans_pca$plot
```

Compare the clusters to the actual species:

```{r kmeans-vs-species}
penguins_kmeans <- res_kmeans_pca$data %>%
  mutate(species = penguins_clean$species)

ggplot(penguins_kmeans, aes(x = species, fill = cluster)) +
  geom_bar(position = "dodge") +
  labs(
    title = "K-means clusters vs true species",
    x = "Species",
    y = "Count",
    fill = "Cluster"
  ) +
  theme_minimal()
```

And a simple confusion table:

```{r kmeans-confusion}
table(penguins_kmeans$species, penguins_kmeans$cluster)
```

## Example 2 – HDBSCAN + UMAP

Now we only change:

* `method = "hdbscan"`
* `dimred = "UMAP"`

Everything else stays the same.

```{r hdbscan-umap}
res_hdbscan_umap <- cluster_and_embed(
  data   = penguins_features,
  method = "hdbscan",
  dimred = "UMAP",
  minPts = 5,
  scale  = TRUE
)

res_hdbscan_umap$plot
```

Compare to species:

```{r hdbscan-vs-species}
penguins_hdbscan <- res_hdbscan_umap$data %>%
  mutate(species = penguins_clean$species)

ggplot(penguins_hdbscan, aes(x = species, fill = cluster)) +
  geom_bar(position = "dodge") +
  labs(
    title = "HDBSCAN clusters vs true species",
    x = "Species",
    y = "Count",
    fill = "Cluster"
  ) +
  theme_minimal()
```

Note: **cluster `0` represents noise** (points not assigned to any cluster).

## Example 3 – Hierarchical clustering (hclust) + UMAP

Hierarchical clustering is often appreciated by biologists because it also gives a **dendrogram**. Here we just use it to assign clusters and visualize them in 2D.

```{r hclust-umap}
res_hclust_umap <- cluster_and_embed(
  data   = penguins_features,
  method = "hclust",
  dimred = "UMAP",
  k      = 3,          # cut the tree into 3 clusters
  scale  = TRUE
)

res_hclust_umap$plot
```

And again, compare to species if needed:

```{r hclust-vs-species}
penguins_hclust <- res_hclust_umap$data %>%
  mutate(species = penguins_clean$species)

table(penguins_hclust$species, penguins_hclust$cluster)
```

## How to reuse this on your own data

Imagine you have a feature table `features_df` where:

* Each row = one object (cell, nucleus, sample, spot, …)
* Each numeric column = one feature (intensity, morphology, expression, …)
* Some columns may be metadata (sample ID, condition, patient, …)

You can plug it directly into:

```{r, eval=FALSE}
res <- cluster_and_embed(
  data   = features_df,
  method = "dbscan",   # "kmeans", "hclust", "dbscan", "hdbscan", "pam"
  dimred = "UMAP",     # or "PCA"
  k      = 4,          # for kmeans / hclust / pam
  eps    = 0.5,        # for DBSCAN
  minPts = 10          # for DBSCAN / HDBSCAN
)

res$plot          # quick visual check
res$data          # original data + cluster column
res$model         # raw clustering model
res$embedding     # 2D coordinates with cluster labels
```

All **non-numeric columns** are preserved as metadata in `res$data`, which makes it easy to analyze cluster composition:

```{r, eval=FALSE}
ggplot(res$data, aes(x = condition, fill = cluster)) +
  geom_bar(position = "fill") +
  labs(
    title = "Cluster composition per condition",
    y = "Proportion",
    fill = "Cluster"
  ) +
  theme_minimal()
```

## Takeaways

With a single function, you now have:

* A **unified interface** for several clustering methods
* Automatic, consistent **2D visualization** using PCA or UMAP
* A clean object (`res`) bundling:

  * original data + cluster labels
  * model
  * embedding
  * plot
  
## Conclusion

Wrapping multiple clustering methods and dimensionality reductions into a single R function makes exploratory analysis faster, simpler, and more consistent. Instead of rewriting the same code for each dataset or algorithm, you can quickly switch methods (K-means, HDBSCAN, DBSCAN, etc.) and focus on interpreting the clusters in their biological or experimental context. As always with unsupervised learning, this is a convenience tool, not a magic wand: the quality of the results still depends on your features, preprocessing, and domain knowledge.