---
title: "Importing Data into R"
description: |
  One of the first steps in any data analysis project is importing your data into R. Whether you're dealing with CSV files, Excel spreadsheets, or SQL databases, R has got you covered. Here's how to get started:
theme: theme.css
author:
  - name: Cédric Hassen-Khodja
date: 2023-09-15
categories: [import]
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Importing CSV Files

**Syntax**:

```{r echo=TRUE, eval=FALSE}
read.csv("file_path", header = TRUE, sep = ",")
```

-   `file_path`: Path to the CSV file you wish to import.
-   `header`: Whether the first row contains column names. Default is TRUE.
-   `sep`: The field separator character. Default is a comma.

### Example: To import a CSV file:

```{r echo=TRUE,eval=FALSE}
my_data <- read.csv("./static/sample_data.csv", header = TRUE, sep = ",")
```

## Importing Excel Files

**Syntax**:

```{r echo=TRUE,eval=FALSE}
library(readxl)
read_excel("file_path", sheet = 1)
```

-   `file_path`: Path to the Excel file you wish to import.
-   `sheet`: The sheet number or name to read.

### Example: To import an Excel file:

```{r echo=TRUE,eval=FALSE}
library(readxl)
my_excel_data <- read_excel("./static/sample_data.xlsx", sheet = 1)
```

## Importing SQL Databases

**Syntax**:

```{r echo=TRUE,eval=FALSE}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "db_path")
dbReadTable(con, "table_name")
```

-   `db_path`: Path to the SQL database you wish to connect to.
-   `table_name`: The name of the table you wish to read.


### Example: To connect to a SQL database and read a table:

```{r echo=TRUE,eval=FALSE}
library(DBI)
con <- dbConnect(RSQLite::SQLite(), "./static/sample_data.db")
my_db_data <- dbReadTable(con, "sample_table")
```

# Conclusion

By mastering these methods, you can easily import various types of data into R, setting the stage for your data analysis journey.
