---
title: "HDBSCAN Clustering: Discovering Variable-Density Structures in Unlabeled Data"
description: |
  Learn to apply HDBSCAN clustering to the Palmer Penguins dataset using only R. This step-by-step guide demonstrates how HDBSCAN improves over DBSCAN by automatically identifying clusters of varying densities and handling outliers.
author:
  - name: Cédric Hassen-Khodja
    url: {}
date: 2025-10-09
categories:
  - [Machine Learning]
  - [Unsupervised Learning]
  - [Clustering]
  - [HDBSCAN]
  - [Visualization]
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
required_packages <- c("palmerpenguins", "dplyr", "ggplot2", "dbscan", "factoextra", "gridExtra", "ggfortify")
new_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if(length(new_packages) > 0) install.packages(new_packages)

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(dbscan)
library(factoextra)
library(gridExtra)
library(ggfortify)
```

## Introduction

**HDBSCAN (Hierarchical Density-Based Spatial Clustering of Applications with Noise)** is an extension of DBSCAN that removes the need to manually set `eps`, allows detection of **clusters with varying densities**, and better handles **noise points**.

We’ll explore its usage in R with the `dbscan` package on the Palmer Penguins dataset.

## Data Preparation

```{r data-prep}
data("penguins", package = "palmerpenguins")
penguins_clean <- penguins %>%
  filter(!is.na(bill_length_mm),
         !is.na(bill_depth_mm),
         !is.na(flipper_length_mm),
         !is.na(body_mass_g)) %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

head(penguins_clean)
```

## Feature Scaling

```{r scaling}
penguins_scaled <- scale(penguins_clean)
```

## Running HDBSCAN

```{r hdbscan}
# minPts = minimum number of points in a dense region (often set to 5)
hdbscan_result <- hdbscan(penguins_scaled, minPts = 5)

# Cluster labels (0 = noise)
table(hdbscan_result$cluster)
```

## Add Cluster Info to Dataset

```{r add-cluster-info}
penguins_compare <- penguins %>%
  filter(!is.na(bill_length_mm),
         !is.na(bill_depth_mm),
         !is.na(flipper_length_mm),
         !is.na(body_mass_g)) %>%
  mutate(hdbscan_cluster = factor(hdbscan_result$cluster),
         species = factor(species))
```

## Visualizing HDBSCAN Clusters with PCA

```{r pca-plot}
pca_res <- prcomp(penguins_scaled)
autoplot(pca_res, data = penguins_compare, colour = 'hdbscan_cluster') +
  labs(title = "HDBSCAN Clusters on PCA Projection") +
  theme_minimal()
```

## Comparing Clusters to True Species

```{r cluster-vs-species}
ggplot(penguins_compare, aes(x = species, fill = hdbscan_cluster)) +
  geom_bar(position = "dodge") +
  labs(title = "HDBSCAN Clusters vs Actual Species",
       x = "Species", y = "Count") +
  theme_minimal()
```

## Confusion Matrix

```{r confusion}
table(penguins_compare$species, penguins_compare$hdbscan_cluster)
```

## Interpretation

HDBSCAN identifies clusters of variable density and automatically filters out noise (cluster `0`). However, on the Palmer Penguins dataset, it only detected **2 clusters**, whereas the ground truth contains **3 distinct species**.

### Why?
- Some species (*e.g.*, Adelie and Chinstrap) are likely **too morphometrically close** to be reliably separated.
- Adjusting `minPts` (tested with values from 3 to 15) had **little impact** on the number or quality of clusters.

> 🔍 **Conclusion:** On this dataset, **HDBSCAN does not outperform DBSCAN**, and tends to under-cluster due to overlapping densities. It is **not always better** and should be evaluated case by case.

## When to Use HDBSCAN?

**Use it when:**

- Clusters vary in shape and density.
- You want to **automatically infer the number of clusters**.
- You expect **noise or ambiguous data points**.

**Limitations:**

- Sensitive to local densities: **may under-detect clusters** if they are not well-separated.
- Slower than DBSCAN.
- Can fail to recover known groupings if the features are not well-discriminated.

## Conclusion

HDBSCAN offers a flexible alternative to DBSCAN, particularly useful when the number of clusters is unknown and densities vary. However, it is **not guaranteed to yield better results**, as shown here on the Palmer Penguins dataset where it failed to recover the 3 biological species. Clustering remains an exploratory task: **no single method works best for all datasets**.
