---
title: "DBSCAN Clustering: Identifying Dense Regions un Unlabeled Data"
description: |
  Learn to apply DBSCAN clustering to the Palmer Penguins dataset to identify natural clusters and outliers. This step-by-step guide covers data preparation, parameter tuning, visualization, and interpretation.
author:
  - name: Cédric Hassen-Khodja
    url: {}
date: 2025-07-08
categories:
  - [Machine Learning]
  - [Unsupervised Learning]
  - [Clustering]
  - [DBSCAN]
  - [Visualization]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
required_packages <- c("palmerpenguins", "dplyr", "ggplot2", "dbscan", "factoextra", "gridExtra")
new_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if(length(new_packages) > 0) install.packages(new_packages)
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(dbscan)
library(factoextra)
library(gridExtra)
```

## Introduction

**DBSCAN (Density-Based Spatial Clustering of Applications with Noise)** is a popular clustering algorithm that groups together points closely packed together while marking isolated points as outliers. Unlike K-Means or Hierarchical clustering, DBSCAN doesn't require specifying the number of clusters in advance.

In this guide, we'll explore DBSCAN on the Palmer Penguins dataset.

## Data Preparation

We select four morphometric features, filtering out missing data.

```{r data-prep}
data("penguins", package = "palmerpenguins")
penguins_clean <- penguins %>%
  filter(!is.na(bill_length_mm),
         !is.na(bill_depth_mm),
         !is.na(flipper_length_mm),
         !is.na(body_mass_g)) %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

head(penguins_clean)
```

## Feature Scaling

DBSCAN is sensitive to scale, so we standardize each variable.

```{r scaling}
penguins_scaled <- scale(penguins_clean)
```

## Determining DBSCAN Parameters

DBSCAN has two parameters:

* **eps**: radius around points.
* **minPts**: minimun number of points required to form a dense region.

We use a k-distance plot to choose **eps**, with **minPts** typically set to dimensionality + 1 (here 5).

```{r k-dist-plot}
kNNdistplot(penguins_scaled, k = 5)
abline(h = 0.55, col = "red", lty = 2)
```

Here, we choose `eps = 0.55` based on the "elbow" in the plot.

## Applying DBSCAN

```{r dbscan-cluster}
dbscan_result <- dbscan(penguins_scaled, eps = 0.55, minPts = 5)
table(dbscan_result$cluster)
```

## Visualizing DBSCAN Clusters with PCA

We project the clusters onto 2 principal components.

```{r pva-viz}
fviz_cluster(dbscan_result, data = penguins_scaled,
             stand = FALSE, ellipse = TRUE, show.clust.cent = FALSE,
             geom = "point", ggtheme = theme_minimal(),
             main = "DBSCAN Clusters on PCA Projection")
```

## Comparing DBSCAN Clusters to Species

Let's assess cluster quality by comparing them to actual species labels.

```{r compare-to-species}
penguins_compare <- penguins %>%
  filter(!is.na(bill_length_mm),
         !is.na(bill_depth_mm),
         !is.na(flipper_length_mm),
         !is.na(body_mass_g)) %>%
  mutate(dbscan_cluster = factor(dbscan_result$cluster),
         species = factor(species))

ggplot(penguins_compare, aes(x = species, fill = dbscan_cluster)) +
  geom_bar(position = "dodge") +
  labs(title = "DBSCAN Clusters vs Actual Species",
       x = "Penguin Species", y = "Count") +
  theme_minimal()
```

## Confusion Matrix

```{r confusion-matrix}
confusion_matrix <- table(penguins_compare$species, penguins_compare$dbscan_cluster)
print(confusion_matrix)
```

## Interpretation

DBSCAN identifies clusters based on density, effectively distinguishing regions of similar penguins. Points labeled as `0` represent outliers or ambiguous individuals. The confusion matrix highlights the following:

* Clusters generally align well with the actual penguin species, with clear separation for most individuals.
* However, DBSCAN appears slightly less effective than K-Means or Hierarchical clustering due to its sensitivity to local density variations, resulting in some penguins being incorrectly classified as outliers or placed into less precise clusters.


## When to Use DBSCAN?

* When data has varying cluster densities.
* When handling noisy data and outliers.
* When the number of clusters is unknown or difficult to determine in advance.

**Limitations:**

* Sensitive to choice of parameters.
* Can struggle with very high-dimensional data.

DBSCAN is a powerful, intuitive method for uncovering meaningful clusters without predefining their number.