---
title: "Exploring Correspondence Analysis with R and factoMineR"
description: |
  Dive into Correspondence Analysis (CA) with R to visualize and analyze the relationships between categories of two qualitative variables. Using `factoMineR`, this practical guide will show you how to apply CA to uncover hidden insights in your categorical data.
theme: theme.css
author:
  - name: Cédric Hassen-Khodja
date: 2024-03-08
categories: 
  - [Categorical Data Analysis]
  - [Visualization]
output:
  distill::distill_article:
    self_contained: false
---


```{r package, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
list.of.packages <- c("FactoMineR")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
```

## Introduction

Correspondence Analysis (CA) is a multivariate statistical technique designed to explore relationships between two or more categorical variables. It's particularly effective for visualizing the strength and patterns of associations in a low-dimensional space. In this post, we'll focus on applying CA using the `factoMineR` package in R, a comprehensive tool for multivariate data analysis.

## Setting Up the Environment

To get started, we'll load the necessary R packages for data manipulation, visualization, and performing PCA.

```{r setup, echo=TRUE}
library(FactoMineR)
```

## Generating a Contingency Table

For demonstration purposes, we'll use a built-in dataset in R and create a contingency table.

```{r generate-data}
set.seed(123) # To ensure reproducibility

data("HairEyeColor")
contingency_table <- margin.table(HairEyeColor, margin = c(1, 2))

# Display the contingency table
contingency_table
```

## Performing Correspondence Analysis

Now, let's perform Correspondence Analysis on the contingency table.

```{r perform-pca, echo=TRUE}
ca_result <- CA(contingency_table, graph = FALSE)
```

## Visualizing the Results

Visualization is a key part of Correspondence Analysis. Let's plot the categories of our variables in the CA space.

```{r visualization}
plot.CA(ca_result, col.row = "steelblue", col.col = "darkred", 
        title = "Correspondence Analysis of Hair and Eye Color")
```

## Interpreting the Results

- **Dimension 1** seems to separate hair colors on a spectrum from `Blond` to `Black/Brown`, with `Blond` hair being on one end and `Black` and `Brown` on the other. Eye colors also align along this dimension, with `Blue` eyes associated with `Blond` hair and `Brown` eyes with `Black/Brown` hair.
- **Dimension 2** might capture additional variation not explained by the first dimension, such as the unique association between `Red` hair and `Green` eyes, as indicated by their significant contributions and positioning in relation to this dimension.