---
title: "Visualizing the Iris Dataset: Enhancing SVM Model Interpretation with R"
description: |
  Dive deeper into the Iris dataset with advanced visualization techniques in R. This guide focuses on leveraging plots and graphics to better understand the distribution of data and the performance of Support Vector Machine (SVM) models. Discover how effective visualization can complement your SVM analysis, offering intuitive insights into complex datasets.
theme: theme.css
author:
  - name: Cédric Hassen-Khodja
date: 2024-03-29
categories: 
  - [Machine Learning]
  - [Visualization]
output:
  distill::distill_article:
    self_contained: false
---


```{r package, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
required_packages <- c("e1071", "ggplot2", "GGally", "prettyunits")
new_packages <- required_packages[!required_packages %in% installed.packages()[,"Package"]]
if(length(new_packages) > 0) install.packages(new_packages)
```

## Introduction

In this guide, we explore how data visualization plays a crucial role in understanding and interpreting Support Vector Machine (SVM) models. Using the Iris dataset, we'll demonstrate various visualization techniques to analyze the dataset's features and the results of our SVM model.

## Preparing the data

Ensure all necessary packages are installed and load them.

```{r setup}
library(e1071)
library(ggplot2)
library(GGally)
```

## Visualizing the Iris Dataset

Visualizing the dataset helps in understanding the distribution of data across different classes and features.

```{r visual}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  theme_minimal() +
  ggtitle("Sepal Dimensions in the Iris Dataset")
```

## Pairwise Feature Relationships

To understand the relationships between all pairs of features, we can use the ggpairs function from the GGally package.

```{r pairs}
ggpairs(iris, aes(color = Species))
```

## Visualizing SVM Model Boundaries

t's insightful to visualize how the SVM model classifies the data. Although directly plotting the SVM decision boundary in higher dimensions is challenging, we can focus on two features at a time.

```{r boundaries}
# Subsetting the dataset for visualization
subset_iris <- iris[, c("Sepal.Length", "Sepal.Width", "Species")]
svm_model_vis <- svm(Species ~ ., data = subset_iris, kernel = "linear")

# Create a grid to plot the decision boundaries
plot(svm_model_vis, subset_iris)
```

## Conclusion

Visualizations provide an intuitive understanding of both the dataset and the SVM model's performance. Through effective graphical representations, we can better understand the relationships between features, assess model accuracy, and make more informed decisions in our data science projects. Let’s continue to leverage the power of visualization in enhancing our machine learning models.