---
title: "Enhancing Classification with Boosting Techniques: Classifying Palmer Penguins"
description: |
  Learn how to apply Boosting algorithms to classify penguin species using the Palmer Penguins dataset. This tutorial walks through data preparation, model construction with Gradient Boosting, cross-validation evaluation, performance analysis, and hyperparameter tuning, showcasing the effectiveness of Boosting methods in improving classification accuracy.
author:
  - name: Cédric Hassen-Khodja
    url: {}
date: 2025-01-17
categories:
  - [Machine Learning]
  - [Classification]
  - [Boosting]
  - [Visualization]
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
required_packages <- c("caret", "xgboost", "palmerpenguins", "dplyr", "ggplot2")
new_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if(length(new_packages) > 0) install.packages(new_packages)
library(caret)
library(xgboost)
library(palmerpenguins)
library(dplyr)
library(ggplot2)
```

## Introduction

Boosting algorithms have become a cornerstone in the field of machine learning, known for their ability to enhance model performance by combining weak learners into a strong ensemble. In this tutorial, we will explore how to apply **Gradient Boosting** using the `xgboost` package in R to classify penguin species from the **Palmer Penguins** dataset. We will cover data preparation, model building, evaluation through cross-validation, performance interpretation, and hyperparameter tuning, demonstrating the superior accuracy Boosting techniques can achieve in classification tasks.

## Data Preparation

As with any machine learning task, proper data preparation is crucial. We will start by cleaning the dataset, handling missing values, and encoding categorical variables appropriately for the Boosting model.

```{r data-prep}
data("penguins")
penguins_clean <- penguins %>%
  filter(!is.na(species),
         !is.na(bill_length_mm),
         !is.na(bill_depth_mm),
         !is.na(flipper_length_mm),
         !is.na(body_mass_g),
         !is.na(sex),
         !is.na(island)) %>%
  mutate(
    species = as.factor(species),
    island = as.factor(island),
    sex = as.factor(sex)
  )

#Quick preview of the cleaned data
head(penguins_clean)
```

### Feature Engineering

Boosting algorithms like XGBoost require numerical input. Therefore, we need to convert categorical variables into numerical formats using one-hot encoding.

```{r feature-engineering}
# One-hot encode categorial variables
penguins_numeric <- penguins_clean %>%
  select(-species) %>%
  mutate_if(is.factor, as.numeric)

# Convert species to numerical labels
species_labels <- as.numeric(penguins_clean$species) - 1  # XGBoost requires labels starting from 0

# Prepare the final dataset
dataset <- as.matrix(penguins_numeric)
```

## Building a Gradient Boosting Model with XGBoost

We will utilize the `xgboost` package to construct a Gradient Boosting model. Our target variable is `species`, and our predictors include `island`, `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`, `body_mass_g`, and `sex`.

```{r model-build}
# Set a seed for reproductibility
set.seed(123)

# Define parameters for XGBoost
params <- list(
  objective = "multi:softprob",
  num_class = length(unique(species_labels)),
  eval_metric = "mlogloss"
)

# Convert data into DMatrix format
dtrain <- xgb.DMatrix(data = dataset, label = species_labels)

# Train the initial XGBoost model
xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 100,
  watchlist = list(train = dtrain),
  verbose = 0
)

print(xgb_model)
```

### Initial Interpretation

- **nrounds=100**: The model is trained for 100 bossting iterations.
- **objective="multi:softprob"**: Specifies a multi-class classification task, outputting probabilities for each class.
- **num_class=3**: There are three penguin species to classify.
- **eval_metric="mlogloss"**: The model is evaluated using multi-class logarithmic loss.

## Feature Importance

Understanding which features contribute most to the model's predictions cn provide valable insights. XGBoost offers built-in methods to assess feature importance.

```{r feature-importance}
importance_matrix <- xgb.importance(feature_names = colnames(dataset),
                                    model = xgb_model)
print(importance_matrix)
xgb.plot.importance(importance_matrix, main = "Feature Importance in XGBoost Model (Penguins)")
```

- **Gain**: The improvement in accuracy brought by a feature to the branches it is on. Higher gain indicates more important features.
- **Cover**: The number of observations related to the feature. Higher cover means the feature is used in many splits.
- **Frequency**: The number of times a feature is used in all generated trees.

### Overall Interpretation

- **bill_length_mm** and **flipper_length_mm** are the most influential features, contributing significantly to species classification.
- **body_mass_g** and **sex** have lower importance, indicating they are less critical for distinguishing between species.

## Model Evaluation via Cross-Validation

To ensure the model's robustness and to prevent overfitting, we perform cross-validation using the `caret` package. We will also tune key hyperparameters to optimize model performance.

```{r model-eval}
# Define cross-validation method
fitControl <- trainControl(
  method = "cv",
  number = 5,
  verboseIter = FALSE,
  allowParallel = TRUE
)

# Define a grid of hyperparameters to search
tuneGrid <- expand.grid(
  nrounds = c(100, 200),
  max_depth = c(3, 6, 9),
  eta = c(0.01, 0.1, 0.3),
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

# Train the model using caret with XGBoost
set.seed(123)
xgb_cv <- train(
  x = dataset,
  y = species_labels,
  method = "xgbTree",
  trControl = fitControl,
  tuneGrid = tuneGrid,
  verbose = FALSE
)

print(xgb_cv)
```
### Interpreting the Results

- **RMSE (Root Mean Squared Error)** measures the average magnitude of prediction errors.
- **Rsquared** indicates the proportion of variance explained by the model.
- **MAE (Mean Absolute Error)** represents the average absolute difference between predicted and actual values.

- The learning rate (eta) of 0.10 provided strong performance, with very low RMSE and MAE values.
- The max_depth of 3 provided a relatively simple model that avoided overfitting.
- Performance metrics (RMSE, Rsquared, and MAE) all indicate that the model has excellent predictive power with low error and high explained variance.

This indicates that the chosen model is likely to generalize well to new data with minimal error.

## Optimal Boosting Model

Based on cross-validation results, we select the best hyperparameters and train the final model.

```{r optimal-model}
# Extract the best hyperparameters
best_params <- xgb_cv$bestTune

# Update parameters with the best hyperparameters
params_optimal <- list(
  objective = "multi:softprob",
  num_class = length(unique(species_labels)),
  eval_metric = "mlogloss",
  max_depth = best_params$max_depth,
  eta = best_params$eta,
  gamma = best_params$gamma,
  colsample_bytree = best_params$colsample_bytree,
  min_child_weight = best_params$min_child_weight,
  subsample = best_params$subsample
)

# Train the optimal XGBoost model
set.seed(123)
xgb_optimal <- xgb.train(
  params = params_optimal,
  data = dtrain,
  nrounds = best_params$nrounds,
  watchlist = list(train = dtrain),
  verbose = 0
)

print(xgb_optimal)
```

This final model incorporates the optimal hyperparameters determined through cross-validation, ensuring enhanced performance and generalization.

## Final Evaluation

To assess the performance of the optimal model, we generate predictions and evaluate them using a confusion matrix and additional metrics.

```{r final-eval}
# Generate predictions
pred_probs <- predict(xgb_optimal, dataset)
pred_matrix <- matrix(pred_probs, ncol = length(unique(species_labels)), byrow = TRUE)
predictions <- max.col(pred_matrix) - 1  # Convert probabilities to class labels

# Convert numerical labels back to factor
pred_species <- factor(predictions, levels = 0:(length(unique(species_labels)) - 1), labels = levels(penguins_clean$species))

# Actual species
actual_species <- penguins_clean$species

# Generate confusion matrix
conf_mat <- confusionMatrix(pred_species, actual_species)
print(conf_mat)
```

### Overall Statistics

These metrics provide a comprehensive view of the model's performance:

1. **Accuracy**:
   - Proportion of correct predictions: **100%**.
   - Indicates that the model correctly classifies 100% of the penguin species in the dataset.

2. **Kappa**:
   - **1**, reflecting excellent agreement between predictions and actual classifications beyond chance.

3. **No Information Rate (NIR)**:
   - Represents the accuracy that could be achieved by always predicting the most frequent class (Adelie: 43.84%).
   - The model significantly outperforms this baseline.

4. **P-value**:
   - Indicates that the model's accuracy is significantly better than random chance (**P < 2.2e-16**).

### Confusion Matrix Visualization

```{r confusion-matrix-plot, echo=FALSE, fig.height=6, fig.width=6}
library(caret)
library(ggplot2)

# Convert confusion matrix to dataframe for plotting
conf_df <- as.data.frame(conf_mat$table)
colnames(conf_df) <- c("Reference", "Prediction", "Freq")

ggplot(data = conf_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 6) +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Confusion Matrix for XGBoost Model (Penguins)", x = "Actual Species", y = "Predicted Species") +
  theme_minimal()
```

    The **Gradient Boosting** model implemented using `xgboost` demonstrates exceptional performance in classifying penguin species within the Palmer Penguins dataset. With an overall accuracy of **99.5%** and near-perfect Kappa, the Boosting approach significantly outperforms baseline models and effectively captures the complex relationships between features.