---
title: "Evaluating Clustering Quality in R: Silhouette and Beyond"
description: |
  Learn how to assess clustering quality in R using internal validation metrics such as the silhouette and Calinski–Harabasz indices, and how to combine them with PCA/UMAP visual checks to make biologically meaningful choices on the Palmer Penguins dataset.
author:
  - name: Cédric Hassen-Khodja
    url: {}
date: 2026-01-09
categories:
  - [Machine Learning]
  - [Unsupervised Learning]
  - [Clustering]
  - [Model Evaluation]
  - [Validation Metrics]
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c(
  "palmerpenguins",
  "dplyr",
  "ggplot2",
  "dbscan",
  "uwot",
  "fpc",
  "cluster",
  "clusterSim"
)

new_packages <- required_packages[!required_packages %in% installed.packages()[, "Package"]]
if (length(new_packages) > 0) install.packages(new_packages)

library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(dbscan)
library(uwot)
library(fpc)
library(cluster)
library(clusterSim)

```

When you run clustering on a new dataset, the next question is always:

> *"Are these clusters any good ?"*

Sometmes you're lucky and have **labels** (e.g., known cell types, species, conditions), and you compare predicted clusters to the ground truth. But in many real-life imaging or omics applications, you only have **unlabeled data**.

In this post, we'll look at **practical ways to evaluate clustering quality in R when you don't have labels**, focusing on:

* The **silhouette coefficient**
* A couple of other **internal validation indices**
* Simple **visual checks** you can reuse on your own data

We'll again use the **Palmer Penguins** dataset so this post fits nicely with the previous "one function to rule them all" clustering article.

## Why bother evaluating clusters?

Clustering is inherently **exploratory**. It's very easy to:

* Over-interpret **pretty PCA/UMAP plots**
* Assume that "more clusters = more insight"
* Or trust the default result of a single algorithm

Evaluation metrics help you:

* **Compare different algorithms** (K-means vs HDBSCAN vs hierarchical, ...)
* **Choose the number of clusters** (for methods like K-means or PAM)
* Detect when clustering is basically **noise + wishful thinking**

They don't replace domain knowledge, but they give you a more objective starting point.

## Data and clustering setup

```{r data-prep}
data("penguins", package = "palmerpenguins")

penguins_clean <- penguins %>%
  filter(
    !is.na(bill_length_mm),
    !is.na(bill_depth_mm),
    !is.na(flipper_length_mm),
    !is.na(body_mass_g)
  )

penguins_features <- penguins_clean %>%
  dplyr::select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g)

head(penguins_features)
```

WE'll also create a small helper to **standardize** and get a distance matrix once, since several metrics use it:

```{r helper}
mat <- scale(as.matrix(penguins_features))
dist_mat <- dist(mat)
```


## Silhouette coefficient: intuition and implementation

The **silhouette coefficient** is probably the most popular internal metric. For each point:

* It compares how close the point is to points in its **own cluster**
* Versus how close it is to points in the **nearest other cluster**

The value ranges from **-1 to 1**:

* Close to **1** -> point is well-matched to its own cluster, far from others
* Around **0** -> boundary point between clusters
* Negative -> might be in the **wrong clusters**

In R, you can compute it via `cluster::silhouette()`.

### Example: K-means with different k

```{r}
set.seed(123)
k_values <- 2:6

sil_results <- lapply(k_values, function(k) {
  km <- kmeans(mat, centers = k)
  sil <- silhouette(km$cluster, dist_mat)
  
  data.frame(
    k = k,
    avg_silhouette = mean(sil[, "sil_width"])
  )
})

sil_results <- bind_rows(sil_results)
sil_results
```

You can also visualize how the average silhouette changes with `k`:

```{r}
ggplot(sil_results, aes(x = k, y = avg_silhouette)) +
         geom_line() +
         geom_point() +
         labs(
           title = "Average silhouette width for K-means",
           x = "Number of clusters (k)",
           y = "Average silhouette"
           ) + theme_minimal()
```

Typical rule of thumb:

* Prefer the **highest** average silhouette
* But also check wether the chosen `k` makes **biological sense**

On the Palmer Penguins dataset, the silhouette is maximal for k = 2, but we know there are 3 species. Because Adelie and Chinstrap partly overlap in morphometric space, the silhouette prefers to merge them into a single, well-separated cluster. In practice, k = 3 is therefore more biologically meaningful, even if its average silhouette is slightly lower.

## Beyond silhouette: Calinski-Harabasz and Davies-Bouldin

The silhouette is great, but it's only metric. The `fpc::cluster.stats()` function gives you several indices at once, including:

* **Calinski-Harabasz (CH) index**
  * Higher is better
  * Favors compact, well-separated clusters

* **Davies-Bouldin (DB) index**
  * Lower is better
  * Penalizes clusters that are not well separated
  
We can wrap everything into a small helper:

```{r}
evaluate_clustering <- function(dist_mat, clusters) {
  cs <- fpc::cluster.stats(d = dist_mat, clustering = clusters)
  
  db <- clusterSim::index.DB(
    x          = mat,       # numeric matrix/data.frame (not scaled if you prefer)
    cl         = clusters,
    centrotypes = "centroids"
  )$DB
  
  data.frame(
    n_clusters     = length(unique(clusters[clusters != 0])),
    avg_silhouette = cs$avg.silwidth,
    ch_index       = cs$ch,
    db_index       = db
  )
}
```

> Note: For density-based methods (DBSCAN/HDBSCAN), cluster `0` is treated as **noise**, which is excluded from the cluster count.

## Comparing algorithms on the same dataset

Let's compare several clustering results on the **same penguin features**:

* K-means (`k = 3`)
* Hierarchical clustering (`k = 3`)
* HDBSCAN (automatic number of clusters)

```{r}
set.seed(123)

## 1) K-Means
km3 <- kmeans(mat, centers = 3)
eval_km3 <- evaluate_clustering(dist_mat, km3$cluster) %>%
  mutate(method = "kmeans_k3")

## 2) Hierarchical clustering
hc <- hclust(dist_mat, method = "ward.D2")
hc3_clusters <- cutree(hc, k = 3)
eval_hc3 <- evaluate_clustering(dist_mat,  hc3_clusters) %>%
  mutate(method = "hclust_k3")

## 3) HDBSCAN
hdb <- hdbscan(mat, minPts = 5)
eval_hdb <- evaluate_clustering(dist_mat, hdb$cluster) %>%
  mutate(method = "hdbscan")

## Combine results
evall_all <- bind_rows(eval_km3, eval_hc3, eval_hdb)
evall_all
```

Interpretation:

* **Higher silhouette / CH** and **lower DB** suggest better separation
* HDBSCAN might find **fewer clusters plus noise**, so metrics are not directly comparable to strict partitionings (K-means, hclust)
* Metrics are **guides**, not absolute truths

In practice, we would keep the 3-cluster solution even though its average silhouette is slightly lower: it matches the known species structure and is easier to interpret biologically. This is a good example of why internal indices should be treated as **guides, not absolute truth** — different metrics encode different preferences, and they must be interpreted in the context of the data and the scientific question.


## Visual sanity checks still matter

Even with several indices, you should always look at the data:

1. **2D embedding plots (PCA or UMAP)**
2. **Cluster composition across conditions or species**
3. **Distributions of key features by cluster**

Example: UMAP colored by K-means clusters vs HDBSCAN clusters.

```{r}
set.seed(123)
umap_coords <- umap(mat, n_neighbors = 15, min_dist = 0.1)

umap_df <- as.data.frame(umap_coords)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df <- umap_df %>%
  mutate(
    km3_cluster = factor(km3$cluster),
    hdb_cluster = factor(hdb$cluster),
    species = penguins_clean$species
  )

#K-Means clusters
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = km3_cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "UMAP embedding colored by K-Means (k = 3)",
    color = "cluster"
  ) +
  theme_minimal()

#HDBSCAN clusters
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = hdb_cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "UMAP embedding colored by HDBSCAN clusters",
    color = "species"
  ) +
  theme_minimal()
```

You can also quickly relate clusters to a **known label** (when available), just to see if the structure aligns with something biological:

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "UMAP embedding colored by true species",
    color = "Species"
  ) +
  theme_minimal()
```

Even when you don't have labels, the same kind of plot is useful to check whether clusters:

* Overlap strongly
* Are dominated by a single dense blob
* Or show clear, interpretable separation

## Conclusion

Internal validation metrics like **silhouette**, **Calinski-Harabasz**, and **Davies-Bouldin** indices give you a more objective way to compare clustering results, especially when you don't have ground truth lables. The help you pick reasonable parameters and avoid over-interpreting arbitrary partitions.

However, they are **guides, not judges**. The best practice is to combine:

* A few **quantitative metrics**,
* Clear **visualizations** (PCA/UMAP, feature distributions),
* And your **domain knowledge** about what patterns are biologically plausible.